---
title: "Geographic Distances and Re-Weighting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{geographic-distances-and-re-weighting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(TARDIS)
```

## Geographic distances

Any geographic raster will have an associated coordinate system, enabling distances between locations in that space to be calculated. Like in the R package `gdistance`, `build_tardis()` uses this information to calculate horizonal distances in metres between adjacent grid cells in a `geoglist` layer, whether in a rectangular or hexagonal grid. `TARDIS` offers a couple of functions to visualise the resulting cost surface, either between grid cells or relative to a single starting point. We will start by building a global, unmasked `tardis` object at 1-degree resolution.

```{r}
# build a global raster with a constant cell value
rst <- rast(nrow = 180, ncol = 360, vals = 1)

# convert to a geoglist, then a tardis object
rtd <- build_tardis(rast_to_geoglist(rst))
```

We can now visualise the average cost of movement through each grid cell in this landscape with the function `cost_surface()`. This returns a `geoglist` which we can plot using the standard method `TARDIS` provides:

```{r}
# calculate cost surface and plot
cs <- cost_surface(rtd)
plot(rtd)
```

This cost surface of geographic distance clearly shows the latitudinal distortion which rectangular grids incur at global scales, with much smaller distances between grid cells towards the poles compared to the equator.

We can also create a surface describing the cost required to reach all cells in `geoglist` layer from a given starting point, using the function `cumulative_cost()`. Any `TARDIS` function which requires points for analysis requires those points to be checked and formatted using another function, `point_check()`, which ensures that all points can be resolved to unmasked cells. In this case, all cells are unmasked, but we still need to use the function to format our input point. We will choose the intersection of the equator and prime meridian - lon-lat [0,0]:

```{r}
# we supply our point as a two column, lon-lat matrix
orig <- point_check(rtd, rbind(0, 0))

# create a cumulative cost surface from this origin point
cc <- cumulative_cost(rtd, origin = orig)

# plot the surface
plot(cc)
```

This is again the output that we expect for a global (spherical) surface. Costs increase radially from our origin point, with distortion of that radius towards the poles, and with the furthest points lying on the intersection of the equator and antimeridian on the opposite side of the globe.

By default, `build_tardis()` will also assume that the input `geoglist` contains bathymetric and/or topographic information in metres. Differences in elevation between adjacent grid cells are consequently used to calculate the addition distance incurred during movement across the grid. At global scales this will be a second order effect, but will matter much more for smaller geographic areas.

## The impact of masking and linking

The above example was a very simple one - a smooth, isotropic sphere, i.e., no additional effects of elevational change on distance and freedom of movement in all directions to all cells in geographic space. `TARDIS` is built for representing connectivity and distance in much more complex landscapes. Let's revisit the Galapagos dataset from previous vignettes. As before, we will select the modern layer, mask it to the land surface and construct our `geoglist` and `tardis` object:

```{r}
# load and mask data
gal <- galapagos()[[5]]
gal_l <- classify(gal, matrix(c(-Inf, 0, NA, 0, Inf, 1), ncol = 3, byrow = T), right = F)

# build the geoglist and tardis
rst <- rast_to_geoglist(gal, gal_l)
rtd <- build_tardis(rst)
```

Lets derive the cost and cumulative cost surfaces as before. For the latter, we will pick a point on the most northerly island, Pinta:

```{r}
# resolve our land point and get our surface plots
pt <- point_check(rtd, rbind(c(-90.75, 0.58)))
cc <- cumulative_cost(rtd, origin = pt)
cs <- cost_surface(rtd)
plot(cs)
plot(cc)
```

Even at this scale, grid cells are still around 1 to 1.5 kilometres apart, compared to vertical differences of several to tens of metres, so the cost surface still looks relatively flat. Even so, we can start to see the impact of topography at this scale, with the volcanic caldera in the north of the largest island, Isabela. Something seems wrong for the other map. Most land is missing, but this is because we did not add in any island links, so our cumulative cost calculations relative to our origin point are completely restricted to Pinta island.

To see how linkage affects distance, we will create two different `tardis` models, one with minimum spanning tree (MST) linkage, the other with Voronoi linkage and visualise the cumulative cost surfaces. Note that links can be added to a `geoglist` even if already contains links - `TARDIS` will automatically delete duplicate links if present:

```{r}
# mst
rst <- link_islands(rst, klink = 1)
rtd1 <- build_tardis(rst)
cc1 <- cumulative_cost(rtd1, origin = pt)

# voronoi
rst2 <- link_islands(rst, klink = NULL)
rtd2 <- build_tardis(rst2)
cc2 <- cumulative_cost(rtd2, origin = pt)

# plot solutions together
par(mfrow = c(1, 2))
plot(cc1, pal = sf.colors(100), breaks = 20)
plot(rst$links, add = T)
plot(cc2, pal = sf.colors(100), breaks = 20)
plot(rst2$links, add = T)
```

Firstly, we can see that the largest distances in the cumulative cost surface have dropped by nearly 250 kilometres. While this affects legend comparison between plots, we can also see that the cost to get to the northern coast of the largest island, Isabela, has dropped from around 200 kilometres to around 5000 kilometres. Clearly linkage structure has a big impact on geographic distance in our landscapes.

## Reweighting with geographic distances

```{r}
rtd <- weight_tardis(rtd, name = "elev10",
                     wfun = function(origin, dest, lnum, ...) {
                       (origin$hdist^2 + abs(origin$vdist * 100)^2)
                     }
                     
                     
)

rtd <- weight_tardis(rtd1, name = "elev10",
                     wfun = function(origin, dest, lnum, ...) {
                      (origin$hdist^2 + abs(origin$vdist * 10)^2)
                     }
)
```
